Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-05-22 document/ms.tex
making plots
some comments from @davidwhogg

An impressive number of exoplanets have been discovered and it is of great
scientific interest to study what we can say about the exoplanet occurrence
rate function given these discoveries.
Every catalog is censored by non-trivial selection effects and detection
efficiencies, and the measurements of the system properties (period, radius,
\etc) are noisy.
We present a general hierarchical probabilistic framework for making justified
inferences about the occurrence rate of exoplanets taking into account survey
completeness \emph{and observational uncertainties}.
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the underlying occurrence rate than the standard
inverse-detection-efficiency weighting procedure.
When applied to an existing catalog of small planet candidates around G dwarfs
(Petigura \etal\ 2013), our method confirms previous measurements of a
flattening radius distribution near Earth's radius.
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat larger.
Under strong assumptions, it has been shown that these populations permit an
uncertain extrapolated estimate of the probability or rate at which Sun-like
stars host true Earth analogs.
We relax these assumptions by modeling the occurrence rate as a
multidimensional Gaussian process and find that the rate density of Earth
analogs is about two percent (per natural logarithmic bin in period and
radius) at low signal-to-noise.

@@

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number density of
planets per natural logarithmic bin in period and radius evaluated at the
location of Earth}
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat and extrapolated from
their sample.
We relax this assumption and extrapolate by only assuming that the
occurrence rate function is ``smooth'' in period and radius.
We enforce this constraint by applying a flexible Gaussian process
regularization to the bin heights.

@@

As well as being a very flexible prior, this Gaussian process model is also
appealing because it is well studied in the machine learning literature.
In particular, there is a very efficient algorithm---elliptical slice sampling
(ESS; \citealt{ess})---for generating samples from the posterior probability
function in \eq{posterior}.
As mentioned above, we actually go one step further and assume uniform
(hyper-)priors on $\ln\mean$ and $\ln\smoothpars$ in order to sample from the
joint posterior
\begin{eqnarray}
p(\ratepars,\,\mean,\,\smoothpars\,|\,\{\data_k\})
\end{eqnarray}
using Algorithm 1 from \citet{ess-hyper}.

@@

@@

@@

@@

% We have made the first measurement of the Earth-like exoplanet population that
% is based on forward modeling, takes proper account of the observational
% uncertainties on planet radii, and makes use of extremely flexible
% non-separable distributions for period and radius.
% We find \hoggtodo{BLAH BLAH BLAH.}

% Although our method makes weaker assumptions than any previous work in this
% area, of course it \emph{does} make strong assumptions.
% It assumes that the data set (from \citealt{petigura}) contains planets and
% only planets (no false positives).
% It assumes that the individual-planet radius uncertainties are also correctly
% determined.
% It assumes that the completeness calculation (again from \citealt{petigura})
% is accurate and smoothly varying with period and radius; it makes no attempt
% to adjust the completeness star-by-star for brightness or variability.
% It assumes that the exoplanets are all independently drawn from the same
% distribution function, with no regard to age, metallicity, or multiplicity.

% Although we have taken account of the radius uncertainties in a principled
% way, we did make one important approximation: We ignored planets with central
% values far outside the period and radius box in which we built the model.
% That is, we assumed that there are no very large outliers in radius
% determination.
% This assumption is a complex one; it involves the effective, end-to-end error
% distribution in the data set, but also smoothness of the distribution function
% at or near the edges of the analysis box.

% If our inferences are right about true Earth analogs, then we expect that
% there are about XXX transiting Earth analogs hidden in the \kepler\ data set.
% We would be able to find some of these Earth analogs if the effective noise
% level in the data could be reduced by a factor of about XXX.
% We have identified a number of directions along which this noise reduction
% might happen, including but not limited to retrospective data-driven or
% physics-driven recalibrations of the \kepler\ focal plane
% (\citealt{hoggwhitepaper}), new methods for performing aperture photometry,
% and better models for stochastic instrument variability or stochastic
% intrinsic stellar variability that take account of frequency structure
% (\citealt{brewer, carter, roberts}).
% If any (or all) of these deliver substantial improvements for \kepler\
% archival data reanalyses, it's Nobel Prize time!  (Or at least \textsl{The
% Colbert Report}?)

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
\ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research builds on collaborations between astronomers and statisticians
forged during a three week workshop on ``Modern Statistical and Computational
Methods for Analysis of Kepler Data'' at SAMSI in June 2013.
This research made use of the NASA \project{Astrophysics Data System}.

>>>>

An impressive number of exoplanets have been discovered and it is of great
scientific interest to study what we can say about the exoplanet occurrence
rate function given these discoveries.
Every catalog is censored by non-trivial selection effects and detection
efficiencies, and the measurements of the system properties (period, radius,
\etc) are noisy.
We present a general hierarchical probabilistic framework for making justified
inferences about the occurrence rate of exoplanets taking into account survey
completeness \emph{and observational uncertainties}.
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the underlying occurrence rate than the standard
inverse-detection-efficiency weighting procedure.
When applied to an existing catalog of small planet candidates around G dwarfs
(Petigura \etal\ 2013), our method confirms previous measurements of a
flattening radius distribution near Earth's radius.
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat larger.
Under strong assumptions, these data have been used to make an extrapolated
estimate of the probability or rate at which Sun-like stars host Earth
analogs.
We relax these assumptions, only requiring that the occurrence rate be a
smooth function of period and radius (using a Gaussian process) and find that
the rate of Earth analogs per star is about 0.02 (per natural logarithmic bin
in period and radius) with large uncertainty.

@@

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number of planets
per star per natural logarithmic bin in period and radius, evaluated at the
period and radius of Earth}
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat, obtaining
$\gammaearth \approx 0.12$.
We relax this assumption and extrapolate by only assuming that the occurrence
rate is a smooth function of period and radius; we find a much lower value for
\gammaearth.
We enforce the smoothness constraint by applying a flexible Gaussian process
regularization to the bin heights.

\dfmtodo{Justify our choice of definition. Not habitability.}

@@

As well as being a very flexible prior, this Gaussian process model is also
appealing because it is well studied in the machine learning literature.
In particular, there is a very efficient algorithm---elliptical slice sampling
(ESS; \citealt{ess})---for generating samples from the posterior probability
function in \eq{posterior}.
Our implementation is based on Jo Bovy's BSD licensed ESS
code\footnote{\url{https://github.com/jobovy/bovy_mcmc/blob/master/bovy_mcmc/%
elliptical_slice.py}}.
As mentioned above, we actually go one step further and assume uniform
(hyper-)priors on $\ln\mean$ and $\ln\smoothpars$ in order to sample from the
joint posterior
\begin{eqnarray}
p(\ratepars,\,\mean,\,\smoothpars\,|\,\{\data_k\})
\end{eqnarray}
using Algorithm 1 from \citet{ess-hyper}.

@@

\dfmtodo{Justify our choice of definition. Not habitability.}

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/results/period.pdf}
\end{center}
\caption{%
The period distribution.
\figlabel{period}}
\end{figure}

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/results/radius.pdf}
\end{center}
\caption{%
The radius distribution.
\figlabel{radius}}
\end{figure}

@@

Probabilistic framework justified by MATHS.

Disagrees in detail with Petigura esp.\ extrapolation.

Agrees qualitatively with Petigura, Dong, Howard, \etc\ but hard to make
quantitative comparisons to other ones (pipeline, stars, \etc).

Why did we use the definition of \gammaearth?

\dfmtodo{%
Predicted number of observable Earth analogs in the \kepler\ dataset.
Low but not zero.
}

@@

Under these assumptions and the dataset, \etc, we stand by this analysis but
really this is a model of the planets in Kepler convolved with any latent
effects of the TERRA pipeline\ldots

A major effect that isn't taken into account is the highest signal-to-noise
blah blah blah\ldots our number is prolly low.

\dfmtodo{%
{\bf Advice:} \quad
Remind us that we want a representation of the full posterior (samples)
because that would let us do better things with the completeness (modeled as
r/R, \etc).
Must explicitly include priors too.
}

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Jo Bovy (IAS),
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research builds on collaborations between astronomers and statisticians
forged during a three week workshop on ``Modern Statistical and Computational
Methods for Analysis of Kepler Data'' at SAMSI in June 2013.
This research made use of the NASA \project{Astrophysics Data System}.
